---
title: "Practicals GWAS. Association analysis and GWAS"
author: "Sodbo Sharapov"
date: "January 22, 2019"
output:
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Outline of the practicals

The practical session consists of three steps:

 1. Exploring phenotype data
 2. Exploring genotype data and quality control
 4. Association analysis and GWAS

## Loading QCed ge03d2clean data

Load GenABEL package
 
```{r}
library(GenABEL)
```

Load phenotype and genotype data

```{r}

load('data/ge02d2clean.RData')


```

Let's check number of IDs and SNPs

```{r}
nids(ge03d2clean)

nsnps(ge03d2clean)
```

## Single SNP association analysis

Let's run association analysis for a single SNP.
For this we will store genotypes for this SNP in separate object.

```{r}

snpnames(ge03d2clean)[1:10]

my_snp_num <- as.numeric(gtdata(ge03d2clean[,'rs1646456']))

my_snp_cha <- as.character(gtdata(ge03d2clean[,'rs1646456']))

```

Let's check the distribution of genotypes across samples

```{r}

table(my_snp_cha)
table(my_snp_num)
```

Let's run logistic regresion with SNP as predictor and dm2 as outcome.

```{r}
log_reg_snp <- glm(phdata(ge03d2clean)$dm2~my_snp_num)

summary(log_reg_snp)
```

As we can see, there is no association at all.
Let's check the distribution of cases and controls across genotypes.

```{r}
table(phdata(ge03d2clean)$dm2,my_snp_cha)
```

Indeed, the distribution of cases and controls across genotypes is random.

## Genome-wide association scan

Now let's run tiny GWAS :). We will use qtscore function of GenABEL package.
Basically, we will run 7000 regressions.

```{r}
qt_simple <- qtscore(dm2, ge03d2clean, trait="binomial")

descriptives.scan(qt_simple , sort="Pc1df")

```

```{r}
qt_sex_age_bmi <- qtscore(dm2 ~ sex + age + bmi, ge03d2clean, trait="binomial")

descriptives.scan(qt_sex_age_bmi , sort="Pc1df")

```

## Manhattan plot

## Regional assoc plot

## Table of genotypes and c/c status for the best SNP

## Saving summary stats

## Additional literature: 

Aulchenko, Yurii S., Karssen, Lennart C., & The GenABEL project developers. (2015). The GenABEL Tutorial. Zenodo. http://doi.org/10.5281/zenodo.19738

Anderson, C. A., Pettersson, F. H., Clarke, G. M., Cardon, L. R., Morris, A. P., & Zondervan, K. T. (2010). Data quality control in genetic case-control association studies. Nature Protocols, 5(9), 1564–73. http://doi.org/10.1038/nprot.2010.116
## Saving data