---
title: "Practicals GWAS. Exploring genotype data"
author: "Sodbo Sharapov"
date: "January 22, 2019"
output:
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Outline of the practicals

The practical session consists of five steps:

 1. Exploring phenotype data
 2. Exploring genotype data
 3. Quality control of genotype data
 4. Association analysis and GWAS
 5. Vizualization of the results

## Loading genotype data

Load GenABEL package
 
```{r}
library(GenABEL)
```

Load phenotype and genotype data

```{r}

data("ge03d2")


```

Let's get the Structure of ge03d2 object

```{r}
str(ge03d2)
```

## Summary of genotype data

Let's have a look at the summary of the genotype data.

How many IDs and SNPs do we have in our data?

```{r}
nids(ge03d2)
nsnps(ge03d2)
```

What information about SNPs do we have?

```{r}
snp_info <- summary(gtdata(ge03d2))

head(snp_info)
```

Description of columns:
  
  1. ...
  2. ...
  3. ...

## Summary for a single SNP

Let's extract information about particular SNP

```{r}

snp_info['rs6212914',]
```

What is the distribution of genotypes for this SNP across samples

```{r}

table(as.character(gtdata(ge03d2[,'rs6212914'])))

```

Let's estiamte effective allele frequency for this SNP using different ways

```{r}

(snp_info['rs6212914','P.12'] + snp_info['rs6212914','P.22'] * 2) / snp_info['rs6212914','NoMeasured']

mean(as.numeric(gtdata(ge03d2[,'rs6212914'])), na.rm=TRUE)

```
Let's perform HWE test for this SNP

```{r}
HWE.show(ge03d2[,'rs6212914'])
```