---
title: "Practicals GWAS. Association analysis and GWAS"
author: "Sodbo Sharapov"
date: "January 22, 2019"
output:
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Outline of the practicals

The practical session consists of three steps:

 1. Exploring phenotype data
 2. Exploring genotype data and quality control
 4. Association analysis and GWAS

## Loading QCed ge03d2clean data

Load GenABEL package
 
```{r}
library(GenABEL)
```

Load phenotype and genotype data

```{r}

load('data/ge02d2clean.RData')


```

Let's check number of IDs and SNPs

```{r}
nids(ge03d2clean)

nsnps(ge03d2clean)
```
## Additional literature: 

Aulchenko, Yurii S., Karssen, Lennart C., & The GenABEL project developers. (2015). The GenABEL Tutorial. Zenodo. http://doi.org/10.5281/zenodo.19738

Anderson, C. A., Pettersson, F. H., Clarke, G. M., Cardon, L. R., Morris, A. P., & Zondervan, K. T. (2010). Data quality control in genetic case-control association studies. Nature Protocols, 5(9), 1564–73. http://doi.org/10.1038/nprot.2010.116
## Saving data